Local Type Inference for Polarised System F with Existentials

### 2023-10-09 oopsla24/tex/02-overview.tex
WIP

\subsection{The Language}

@@

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }

    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
  \caption{Declarative Types of \fexists}
  \label{fig:declarative-types}
\end{figure}

@@

>>>>

\subsection{What types do we infer?}

\subsection{The Language of Types}

@@

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }
    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
  \caption{Declarative Types of \fexists}
  \label{fig:declarative-types}
\end{figure}

@@

\subsection{The Language of Terms}

In \cref{fig:declarative-terms}, we define the the language of terms of 
\fexists. The language combines \systemf with the \CBPV approach.

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a (positive) term variable; 
        \ilyam{why no negatives?}
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ operate similarly to 
        the bind of a monad: they take a suspended computation $[[v]]$,
        apply it to a list of arguments, bind the result 
        (which is expected to be pure) to a variable $[[x]]$,
        and continue with a computation $[[c]]$.
        If the resulting type of the application is unique, 
        one can omit the type annotation, as in the second form:
        it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

\paragraph{Missing constructors}
Notice that the language does not have first-class applications: 
their role is played by the applicative let forms, binding 
the result of a \emph{fully applied} function to a variable.
Also notice that the language does not have a type application 
(i.e. the eliminator of $[[∀]]$) and dually, 
it does not have \pack (i.e. the constructor of $[[∃]]$).
This is because the instantiation of polymorphic and existential 
types is inferred by the algorithm.  \ilyam{refer to the extension chapter}

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottc\ottinterrule
    }

    \ottgrammartabular{
      \ottv\ottinterrule
    }
  \end{multicols}
  \caption{Declarative Terms of \fexists}
  \label{fig:declarative-terms}
\end{figure}

\subsection{The key ideas of the algorithm}

### 2023-10-09 oopsla24/tex/03-declarative-system.tex

\subsection{The Language}

\begin{definition}[Declarative Types]
  \hfill
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }

    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
\end{definition}

>>>>

The declarative system serves as a specification of the
type inference algorithm. It consists of two main parts:
the subtyping and the type inference.

\subsection{Subtyping}
It is represented by a set of inference rules shown in
\cref{fig:declarative-subtyping}.

\begin{figure}[h]
  \begin{multicols}{2}
    \ottdefnDOneNsub{}

    \ottdefnDOnePsup{}
  \end{multicols}
  \hfill\\
  \begin{multicols}{2}
    \ottdefnDOneNeq{}

    \ottdefnDOnePeq{}
  \end{multicols}
  \caption{Declarative Subtyping}
  \label{fig:declarative-subtyping}
\end{figure}

\paragraph{Invariance of Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants makes the system equivalent to \systemf, 
and thus, undecidable. 
However, \ruleref{\ottdruleDOneShiftDLabel} might be
relaxed to a covariant form, as we will discuss in \cref{todo}.

As discussed in \cref{todo}, 
the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} are the only non-algorithmic ones.
For convenience of representation, we compose the left-hand side rule and
the right-hand side rule into one, and use substitution $[[σ]]$ to 
represent instantiation.

The substitution application is defined 
in a standard way, avoiding capture of bound variables,
and preserving the variables that are out of the substitution domain.
The domain and the range of the substitutions are
specified by notation $[[Γ2 ⊢ σ : Γ1]]$.
For instance, the notation $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.

\subsection{Type Inference}

\begin{figure}[h]
  \ottdefnDTNInf{}

  \hfill

  \begin{multicols}{2}
  \ottdefnDTPInf{}
  \\
  \ottdefnDTSpinInf{}
  \end{multicols}
  \hfill

  \caption{Declarative Subtyping}
  \label{fig:declarative-inference}
\end{figure}

\subsection{Properties of the Declarative System}

Now, we present selected properties of the declarative system,
which are important for the correctness of the algorithm.

\begin{lemma}
  Variables do not have proper subtypes and supertypes
\end{lemma}

\begin{lemma}
  Free Variable propagation: $[[Γ ⊢ iN ≤ iM]]$ 
  implies $[[fv(iN) ⊆ fv(iM)]]$.
\end{lemma}

\begin{lemma}
  Subtyping is Reflexive and Transitive, and is preserved by substitution
\end{lemma}

\begin{lemma}
  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) ≈ nf(iM)]]$.
\end{lemma}

\begin{lemma}[Characterization of Supertypes]
\end{lemma}
